# Data preparation {#data-preparation}

The staged panel contains 360 geographically dispersed Microflora Danica
samples: 45 from each of eight habitat classes. It uses the same locked sample
and representative genome selections as the
[gifter case study](https://github.com/alberdilab/gifter/tree/f04f7de0691c29874f7fbe17b915990b832e2a4d/manuscript/analysis/r10-mfd).

## Input data

| File | Contents |
| --- | --- |
| [`data/mag_relative_abundance.tsv.xz`](https://github.com/alberdilab/hilldiv3-ms/blob/main/data/mag_relative_abundance.tsv.xz) | Relative abundances for 5,518 representative MAGs (rows) and 360 samples (columns). The first column is `genome_id`. |
| [`data/sample_metadata.tsv`](https://github.com/alberdilab/hilldiv3-ms/blob/main/data/sample_metadata.tsv) | Sample metadata in abundance-column order, including habitat class, ontology, location, date, and selection audit. |
| [`data/representative_metadata.tsv`](https://github.com/alberdilab/hilldiv3-ms/blob/main/data/representative_metadata.tsv) | Metadata for the 95% ANI species representatives in abundance-row order. |
| [`data/representative_tree.nwk`](https://github.com/alberdilab/hilldiv3-ms/blob/main/data/representative_tree.nwk) | Source tree pruned to the 5,518 representative `genome_id` tips, retaining source branch lengths. |

## Load the staged panel

The committed tables and tree are the input to the forthcoming analyses. Load
them from the repository root:

```{r load_staged_data, message=FALSE}
sample_metadata <- data.table::fread(
  "data/sample_metadata.tsv", data.table = FALSE
)
genome_metadata <- data.table::fread(
  "data/representative_metadata.tsv", data.table = FALSE
)
genome_abundance <- data.table::fread(
  cmd = "xz -dc data/mag_relative_abundance.tsv.xz",
  data.table = FALSE, check.names = FALSE
)
genome_tree <- ape::read.tree("data/representative_tree.nwk")
```

The selected panel has 45 samples in each habitat class:

```{r summarize_habitats}
habitats <- table(sample_metadata$habitat_label)
knitr::kable(
  data.frame(habitat = names(habitats), samples = as.integer(habitats)),
  col.names = c("Habitat", "Samples")
)
```

## Check the staged objects

The abundance rows and columns must align with the metadata, and the tree
must contain exactly the representative genomes:

```{r check_staged_data}
stopifnot(
  nrow(sample_metadata) == 360L,
  nrow(genome_metadata) == 5518L,
  nrow(genome_abundance) == 5518L,
  ncol(genome_abundance) == 361L,
  length(habitats) == 8L,
  all(habitats == 45L),
  identical(genome_abundance$genome_id, genome_metadata$genome_id),
  identical(names(genome_abundance)[-1L], sample_metadata$sample_id),
  setequal(genome_tree$tip.label, genome_metadata$genome_id)
)
profile_totals <- colSums(as.matrix(genome_abundance[-1L]))
stopifnot(all(is.finite(profile_totals)),
          all(profile_totals > 0.99 & profile_totals < 1.01))
cat(sprintf(
  "%d samples, %d representative genomes, %d tree tips; profile sums %.6f–%.6f.\n",
  nrow(sample_metadata), nrow(genome_metadata), ape::Ntip(genome_tree),
  min(profile_totals), max(profile_totals)
))
```

The abundance values are **proportions of the published MFD MAG profile**, not
genome or read counts. The preparation script keeps strain (`t__`) rows from
the Sylph profiles, maps MAGs to their published `secondary_cluster`, sums
each 95% ANI cluster, and divides percentages by 100. It assigns each cluster
profile to the deposited representative MAG. It does not renormalize profiles.
The representative abundance therefore stands for its cluster; it does not
imply that all members have identical gene content. These values describe the
MFD MAG profile, not the fraction of all DNA in each sample.

## Sources

The primary study is [Singleton and colleagues, *The Microflora Danica atlas of
Danish environmental microbiomes*](https://doi.org/10.1038/s41586-025-09794-2).
The [Microflora Danica data deposit](https://doi.org/10.5281/zenodo.17162544)
provides `Nitrifiers.tar.gz`, including the published MAG metadata and Sylph
abundance table. Please cite both the study and deposit when using these data,
and consult the deposit for its reuse terms.

The sample and representative tables are copies of the locked panels at
[`gifter` commit `f04f7de`](https://github.com/alberdilab/gifter/tree/f04f7de0691c29874f7fbe17b915990b832e2a4d/manuscript/analysis/r10-mfd).
The source tree is
[`gtdb-shallow.tree`](https://github.com/cmc-aau/mfd_sr_mags/blob/957591d4104462ec6e52ab58d85a69798057e582/analysis/datasets/gtdb-shallow.tree)
from the MFD companion repository at commit `957591d`. Tips outside the
representative panel were removed.

## Rebuild and verify

The committed files are ready to use. To rebuild them from the pinned public
sources, install R packages `data.table` and `ape`, then run:

```sh
bash scripts/fetch-source-data.sh
Rscript scripts/prepare-data.R
Rscript scripts/check-data.R
```

The source archive is about 228 MB. Downloads remain in the ignored
`sources/` directory. The preparation script also accepts paths to existing
MAG metadata, abundance TSV.xz, and source tree files, in that order.

To verify the committed snapshot byte for byte:

```sh
shasum -a 256 -c data/SHA256SUMS
```

The checker verifies sample and genome IDs, 45 samples per habitat, table
ordering, finite nonnegative abundances, near-unit profile sums, and exact
tree-tip identity. In this snapshot, sample profile totals range from
0.999988 to 1.000015.
