# Initial diversity analyses {#diversity}

We use [hilldiv3](https://alberdilab.github.io/hilldiv3/) to describe diversity
in the selected panel. Neutral Hill numbers use representative relative
abundances; phylogenetic Hill numbers also use the pruned representative tree.
We evaluate diversity orders $q=0$ (all detected units receive full weight),
$q=1$ (abundance-weighted diversity), and $q=2$ (common units weighted more
strongly). For neutral diversity, $q=0$ is the count of detected
representatives. Functional diversity awaits a matched trait or gene-function
table.

```{r diversity_setup, include=FALSE}
stopifnot(requireNamespace("hilldiv3", quietly = TRUE),
          identical(rownames(overview_abundance), genome_metadata$genome_id),
          identical(colnames(overview_abundance), sample_metadata$sample_id))

# The installed 3.0.0 release computes one type per hilldiv() call. Separate
# calls also make the neutral and phylogenetic inputs explicit in newer builds.
plain_hill_result <- function(x) {
  class(x) <- "data.frame"
  x
}

alpha_neutral <- plain_hill_result(hilldiv3::hilldiv(
  overview_abundance, q = c(0, 1, 2), type = "neutral"
))
alpha_phylogenetic <- plain_hill_result(hilldiv3::hilldiv(
  overview_abundance, q = c(0, 1, 2), tree = genome_tree,
  type = "phylogenetic", reference = "pool"
))
alpha_neutral$type <- "Neutral"
alpha_phylogenetic$type <- "Phylogenetic"
alpha_diversity <- rbind(alpha_neutral, alpha_phylogenetic)
alpha_diversity$habitat <- sample_metadata$habitat_label[
  match(alpha_diversity$sample, sample_metadata$sample_id)
]
stopifnot(!anyNA(alpha_diversity$habitat),
          all(is.finite(alpha_diversity$value)),
          all(alpha_diversity$value > 0))
alpha_diversity$habitat <- factor(alpha_diversity$habitat,
                                  levels = rev(habitat_levels))
alpha_diversity$type <- factor(alpha_diversity$type,
                               levels = c("Neutral", "Phylogenetic"))
alpha_diversity$order <- factor(alpha_diversity$q, levels = c(0, 1, 2),
                                labels = c("q = 0", "q = 1", "q = 2"))
```

## Within-sample diversity

```{r diversity_alpha_plot, fig.width=11, fig.height=7.5, fig.alt="Boxplots and sample points for neutral and phylogenetic Hill diversity at orders 0, 1, and 2 across eight habitats."}
ggplot2::ggplot(alpha_diversity,
                ggplot2::aes(x = value, y = habitat)) +
  ggplot2::geom_boxplot(outlier.shape = NA, width = 0.55,
                        fill = "#e6c7c7", colour = "#702626") +
  ggplot2::geom_jitter(height = 0.13, width = 0, alpha = 0.28,
                       size = 0.65, colour = "#702626") +
  ggplot2::facet_wrap(ggplot2::vars(type, order), ncol = 3,
                      scales = "free_x") +
  ggplot2::labs(x = "Hill diversity", y = NULL) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
                 strip.text = ggplot2::element_text(face = "bold"))
```

*Hill numbers for 45 samples in each habitat class. Boxes show medians and
interquartile ranges; points are individual samples. Horizontal scales differ
by panel.*

```{r diversity_alpha_table}
q1 <- alpha_diversity[alpha_diversity$q == 1, ]
q1_summary <- aggregate(
  value ~ habitat + type, data = q1,
  FUN = function(x) sprintf("%.1f [%.1f–%.1f]",
                            median(x), unname(quantile(x, 0.25)),
                            unname(quantile(x, 0.75)))
)
q1_neutral <- q1_summary[q1_summary$type == "Neutral", c("habitat", "value")]
q1_phylo <- q1_summary[q1_summary$type == "Phylogenetic", c("habitat", "value")]
names(q1_neutral)[2] <- "Neutral q = 1"
names(q1_phylo)[2] <- "Phylogenetic q = 1"
q1_table <- merge(q1_neutral, q1_phylo, by = "habitat", sort = FALSE)
q1_table <- q1_table[match(habitat_levels, q1_table$habitat), ]
names(q1_table)[1] <- "Habitat"
rownames(q1_table) <- NULL
knitr::kable(q1_table, row.names = FALSE)
```

*Median [interquartile range] of $q=1$ diversity by habitat.*

```{r diversity_alpha_interpretation, echo=FALSE, results='asis'}
neutral_medians <- tapply(q1$value[q1$type == "Neutral"],
                          q1$habitat[q1$type == "Neutral"], median)
cat(sprintf(
  "At $q=1$, median neutral diversity ranges from %.1f in %s to %.1f in %s.\n\n",
  min(neutral_medians), names(which.min(neutral_medians)),
  max(neutral_medians), names(which.max(neutral_medians))
))
```

The phylogenetic values use hilldiv3's common pooled reference tree depth, so
samples are evaluated on the same scale. Their numerical values should be
compared within the phylogenetic panels, rather than interpreted as counts of
MAGs. Neutral $q=0$ is the number of representatives with a positive reported
abundance; it is sensitive to very small values in the source profiles.

## Partitioning across habitats

The next calculation partitions diversity across individual samples, habitat
classes, and the complete panel. At each step, hilldiv3 reports a beta factor:
the effective number of distinct units added at that scale. The two beta
factors multiply to give the total gamma-to-alpha ratio. Each sample has equal
weight; the design has 45 samples in every habitat class.

```{r diversity_partition}
habitat_metadata <- data.frame(
  habitat_class = sample_metadata$habitat_class,
  row.names = sample_metadata$sample_id
)
partition_neutral <- plain_hill_result(hilldiv3::hillpart(
  overview_abundance, q = c(0, 1, 2),
  hierarchy = ~ habitat_class, metadata = habitat_metadata,
  type = "neutral"
))
partition_phylogenetic <- plain_hill_result(hilldiv3::hillpart(
  overview_abundance, q = c(0, 1, 2), tree = genome_tree,
  hierarchy = ~ habitat_class, metadata = habitat_metadata,
  type = "phylogenetic"
))
partition_neutral$type <- "Neutral"
partition_phylogenetic$type <- "Phylogenetic"
partition <- rbind(partition_neutral, partition_phylogenetic)
stopifnot(all(is.finite(partition$diversity)),
          all(is.finite(partition$beta[partition$scale != "sample"])))

within <- partition[partition$scale == "habitat_class",
                    c("type", "q", "beta")]
between <- partition[partition$scale == "total",
                     c("type", "q", "beta", "diversity")]
names(within)[3] <- "Within habitats"
names(between)[3:4] <- c("Between habitats", "Gamma diversity")
partition_table <- merge(within, between, by = c("type", "q"), sort = FALSE)
partition_table <- partition_table[order(match(partition_table$type,
                                                c("Neutral", "Phylogenetic")),
                                          partition_table$q), ]
knitr::kable(partition_table, digits = 2)
```

*Hierarchical Hill-number partition. Beta factors describe sample-to-habitat
and habitat-to-panel turnover; gamma is diversity in the full panel.*

At $q=1$, neutral beta diversity is
`r sprintf("%.2f", partition_table[["Within habitats"]][partition_table$type == "Neutral" & partition_table$q == 1])`
from samples to habitats and
`r sprintf("%.2f", partition_table[["Between habitats"]][partition_table$type == "Neutral" & partition_table$q == 1])`
from habitats to the full panel. The corresponding phylogenetic factors are
`r sprintf("%.2f", partition_table[["Within habitats"]][partition_table$type == "Phylogenetic" & partition_table$q == 1])`
and
`r sprintf("%.2f", partition_table[["Between habitats"]][partition_table$type == "Phylogenetic" & partition_table$q == 1])`.

These are descriptive comparisons of a geographically selected panel. Habitat
and collection year are not independently balanced, so these summaries do not
by themselves establish habitat effects. Relative abundances also cannot
support read-count rarefaction or sampling-depth uncertainty estimates.
