# Pairwise beta diversity {#pairwise-beta}

The [diversity chapter](#diversity) partitions turnover across the whole panel.
Here we compare every pair of the 360 samples with `hilldiv3::hillpair()`.
For each pair, we use the C dissimilarity, the complement of the C overlap
derived from multiplicative Hill beta diversity. It ranges from zero for
identical communities to one for no overlap. We calculate neutral and
phylogenetic values at $q=0$, $q=1$, and $q=2$ using the same representative
profiles and tree as above. Each type and order has
`r choose(ncol(overview_abundance), 2)` unique sample pairs.

```{r pairwise_setup, include=FALSE}
# Explicit sample IDs make the pair-to-habitat mapping unambiguous.
pair_neutral <- plain_hill_result(hilldiv3::hillpair(
  overview_abundance, q = c(0, 1, 2), metric = "C",
  type = "neutral", out = "tibble"
))
pair_phylogenetic <- plain_hill_result(hilldiv3::hillpair(
  overview_abundance, q = c(0, 1, 2), metric = "C",
  tree = genome_tree, type = "phylogenetic", out = "tibble"
))
pair_neutral$type <- "Neutral"
pair_phylogenetic$type <- "Phylogenetic"
pairwise <- rbind(pair_neutral, pair_phylogenetic)
sample_ids <- sample_metadata$sample_id
first_idx <- match(pairwise$first, sample_ids)
second_idx <- match(pairwise$second, sample_ids)
stopifnot(
  nrow(pairwise) == 2L * 3L * choose(length(sample_ids), 2L),
  !anyNA(first_idx), !anyNA(second_idx),
  all(first_idx < second_idx),
  identical(sort(unique(pairwise$q)), c(0, 1, 2)),
  identical(unique(pairwise$metric), "C"),
  all(is.finite(pairwise$value)),
  all(pairwise$value >= -1e-10 & pairwise$value <= 1 + 1e-10)
)
pairwise$habitat_first <- sample_metadata$habitat_label[first_idx]
pairwise$habitat_second <- sample_metadata$habitat_label[second_idx]
pairwise$comparison <- ifelse(
  pairwise$habitat_first == pairwise$habitat_second,
  "Within habitat", "Between habitats"
)
pairwise$type <- factor(pairwise$type,
                        levels = c("Neutral", "Phylogenetic"))
```

## Within and between habitats

The table summarizes all sample pairs in each category. A habitat contributes
990 within-habitat pairs; two distinct habitats contribute 2,025 pairs. These
pairs share samples, so their values are descriptive observations rather than
independent replicates for a significance test.

```{r pairwise_within_between}
pair_summary <- do.call(rbind, lapply(
  split(pairwise, list(pairwise$type, pairwise$q,
                       pairwise$comparison), drop = TRUE),
  function(d) data.frame(
    Type = as.character(d$type[1]), q = d$q[1],
    Comparison = d$comparison[1], Pairs = nrow(d),
    Median = median(d$value),
    Q1 = unname(quantile(d$value, 0.25)),
    Q3 = unname(quantile(d$value, 0.75))
  )
))
pair_summary <- pair_summary[
  order(match(pair_summary$Type, c("Neutral", "Phylogenetic")),
        pair_summary$q,
        match(pair_summary$Comparison,
              c("Within habitat", "Between habitats"))),
]
pair_summary$`Median [IQR]` <- sprintf(
  "%.3f [%.3f–%.3f]", pair_summary$Median,
  pair_summary$Q1, pair_summary$Q3
)
knitr::kable(pair_summary[c("Type", "q", "Comparison", "Pairs",
                            "Median [IQR]")], row.names = FALSE)
```

*C dissimilarity for all unique sample pairs, summarized as median
[interquartile range]. Each within-habitat row contains eight sets of 990
pairs; each between-habitat row contains 28 sets of 2,025 pairs.*

```{r pairwise_within_between_interpretation, echo=FALSE, results='asis'}
q1_pairs <- pair_summary[pair_summary$q == 1, ]
for (diversity_type in c("Neutral", "Phylogenetic")) {
  d <- q1_pairs[q1_pairs$Type == diversity_type, ]
  within_median <- d$Median[d$Comparison == "Within habitat"]
  between_median <- d$Median[d$Comparison == "Between habitats"]
  cat(sprintf(
    "At $q=1$, %s C dissimilarity has median %.3f within habitats and %.3f between habitats.\n\n",
    tolower(diversity_type), within_median, between_median
  ))
}
```

Neutral between-habitat medians are close to one at every order, including
$q=2$, where common representatives carry the most weight. Phylogenetic
between-habitat medians decrease from $q=0$ to $q=2$, consistent with greater
shared branch structure among the more abundant representatives.

## Habitat pair comparisons

We take the median across the sample pairs for each pair of habitats. A
diagonal cell summarizes pairs of *different samples* from the same habitat;
it is not a self-comparison. The heatmaps show $q=1$, with a common zero-to-one
colour scale for both diversity types.

```{r pairwise_habitat_heatmap_data, include=FALSE}
q1_pairwise <- pairwise[pairwise$q == 1, ]
habitat_short <- c(
  "Field soil", "Grassland soil", "Forest soil", "Urban greenspace",
  "Bog/fen soil", "Freshwater sediment", "Saltwater sediment",
  "Wastewater"
)
stopifnot(length(habitat_short) == length(habitat_levels))
habitat_i <- match(q1_pairwise$habitat_first, habitat_levels)
habitat_j <- match(q1_pairwise$habitat_second, habitat_levels)
stopifnot(!anyNA(habitat_i), !anyNA(habitat_j))
habitat_pairs <- aggregate(
  q1_pairwise$value,
  by = list(type = as.character(q1_pairwise$type),
            lower = pmin(habitat_i, habitat_j),
            upper = pmax(habitat_i, habitat_j)),
  FUN = median
)
names(habitat_pairs)[4] <- "median"
heatmap_data <- expand.grid(
  type = c("Neutral", "Phylogenetic"),
  row = seq_along(habitat_levels),
  column = seq_along(habitat_levels),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
pair_key <- function(type, i, j) paste(type, pmin(i, j), pmax(i, j))
heatmap_data$median <- habitat_pairs$median[
  match(pair_key(heatmap_data$type, heatmap_data$row,
                 heatmap_data$column),
        pair_key(habitat_pairs$type, habitat_pairs$lower,
                 habitat_pairs$upper))
]
stopifnot(nrow(habitat_pairs) == 2L * 36L,
          !anyNA(heatmap_data$median))
heatmap_data$Habitat1 <- factor(
  habitat_short[heatmap_data$row], levels = rev(habitat_short)
)
heatmap_data$Habitat2 <- factor(
  habitat_short[heatmap_data$column], levels = habitat_short
)
```

```{r pairwise_habitat_heatmap, fig.width=9.5, fig.height=11, fig.alt="Two heatmaps of median order-one C dissimilarity for all pairs of eight habitats, for neutral and phylogenetic diversity."}
ggplot2::ggplot(heatmap_data,
                ggplot2::aes(x = Habitat2, y = Habitat1, fill = median)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(ggplot2::aes(label = sprintf("%.2f", median),
                                 colour = median > 0.55), size = 3) +
  ggplot2::scale_colour_manual(values = c(`FALSE` = "#292929",
                                          `TRUE` = "white"), guide = "none") +
  ggplot2::scale_fill_gradient(low = "#f7f7f7", high = "#8b1e1e",
                               limits = c(0, 1), name = "C dissimilarity") +
  ggplot2::facet_wrap(~ type, ncol = 1) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(panel.grid = ggplot2::element_blank(),
                 axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
                 strip.text = ggplot2::element_text(face = "bold"))
```

*Median $q=1$ C dissimilarity across 990 pairs on each diagonal and 2,025
pairs in each off-diagonal habitat combination. The matrices are symmetric.*

Saltwater sediment and wastewater are nearly maximally dissimilar to the soil
habitats at $q=1$ in the neutral analysis. Their phylogenetic dissimilarities
are lower, showing that the sample pairs retain some shared evolutionary
branch structure despite little overlap in representative profiles.

## Sample-level ordination

Principal coordinates analysis (PCoA) projects the $q=1$ dissimilarities into
two dimensions. We apply a Cailliez correction because C dissimilarity need
not be Euclidean. Axis percentages refer to the corrected distance matrix;
the plot is a two-dimensional summary of the complete pairwise matrix.

```{r pairwise_ordination_data, include=FALSE}
ordination_rows <- lapply(c("Neutral", "Phylogenetic"), function(diversity_type) {
  d <- q1_pairwise[q1_pairwise$type == diversity_type, ]
  first <- match(d$first, sample_ids)
  second <- match(d$second, sample_ids)
  stopifnot(nrow(d) == choose(length(sample_ids), 2L),
            !anyDuplicated(paste(first, second)),
            all(first < second))
  distances <- matrix(0, nrow = length(sample_ids),
                      ncol = length(sample_ids),
                      dimnames = list(sample_ids, sample_ids))
  distances[cbind(first, second)] <- d$value
  distances[cbind(second, first)] <- d$value
  ord <- stats::cmdscale(stats::as.dist(distances), k = 2,
                         eig = TRUE, add = TRUE)
  variance <- 100 * ord$eig[1:2] / sum(pmax(ord$eig, 0))
  data.frame(
    sample_id = sample_ids,
    habitat = sample_metadata$habitat_label,
    PCoA1 = ord$points[, 1], PCoA2 = ord$points[, 2],
    panel = sprintf("%s (axes: %.1f%%, %.1f%%)",
                    diversity_type, variance[1], variance[2])
  )
})
ordination <- do.call(rbind, ordination_rows)
ordination$habitat <- factor(ordination$habitat, levels = habitat_levels)
stopifnot(nrow(ordination) == 2L * length(sample_ids),
          all(is.finite(ordination$PCoA1)),
          all(is.finite(ordination$PCoA2)))
```

```{r pairwise_ordination, fig.width=11, fig.height=5.5, fig.alt="Principal coordinates plots of order-one neutral and phylogenetic C dissimilarity among 360 samples, coloured by habitat."}
ggplot2::ggplot(ordination,
                ggplot2::aes(x = PCoA1, y = PCoA2, colour = habitat)) +
  ggplot2::geom_point(alpha = 0.75, size = 1.5) +
  ggplot2::scale_colour_manual(values = grDevices::hcl.colors(
    length(habitat_levels), palette = "Dark 3")) +
  ggplot2::facet_wrap(~ panel, scales = "free") +
  ggplot2::labs(x = "PCoA 1", y = "PCoA 2", colour = "Habitat") +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
                 strip.text = ggplot2::element_text(face = "bold"))
```

*Each point is one sample. Axis percentages are the shares of positive
eigenvalues after the Cailliez correction; coordinates are comparable only
within a panel.*

The pairwise patterns describe this selected panel. Pairs sharing samples are
dependent, and habitat class is partly confounded with sampling year. We do
not treat visual separation or differences between pairwise medians as a
formal test of a habitat effect. As elsewhere, $q=0$ responds to very small
reported abundances in the source profiles.
