# Taxonomic shares and dissimilarity contributions {#shares}

The [diversity](#diversity) and [pairwise beta diversity](#pairwise-beta)
chapters describe whole communities. Here we use `hilldiv3::hillshare()` to
allocate each sample's Hill **power sum** to sets of representative MAGs.
A share is a fraction between zero and one, not a fraction of the effective
Hill number reported by `hilldiv()`. The calculation is descriptive; it does
not test whether a phylum is enriched in a habitat.

This chapter was rendered with `hilldiv3` version
`r as.character(utils::packageVersion("hilldiv3"))`.

## Define the genome sets

We use four phyla that have substantial abundance or dissimilarity
contributions in the field–grassland soil comparison. The `Other` set contains
every remaining representative, so the five sets are disjoint and exhaustive.
Each representative stands for its published 95% ANI cluster; the shares
describe those cluster profiles, not the abundance of every MAG within a
cluster.

```{r share_sets}
phylum <- sub("^p__", "", genome_metadata$phylum)
focus_phyla <- c("Pseudomonadota", "Actinomycetota",
                  "Acidobacteriota", "Thermoproteota")
contrast_habitats <- c("Agricultural field soil", "Natural grassland soil")
share_sets <- lapply(focus_phyla, function(name) {
  genome_metadata$genome_id[phylum == name]
})
names(share_sets) <- focus_phyla
share_sets$Other <- genome_metadata$genome_id[!phylum %in% focus_phyla]
stopifnot(sum(lengths(share_sets)) == nrow(genome_metadata),
          setequal(unlist(share_sets, use.names = FALSE),
                   genome_metadata$genome_id))
```

`hillshare()` returns one row for each sample, order, diversity type, and set.
The phylogenetic call uses the same representative tree as the earlier
chapters. Both diversity types are requested together.

```{r share_hillshare, message=FALSE, class.source='script-source hilldiv3-source'}
shares <- hilldiv3::hillshare(
  overview_abundance, q = c(0, 1, 2), tree = genome_tree,
  type = c("neutral", "phylogenetic"), sets = share_sets
)
```

```{r share_checks, include=FALSE}
stopifnot(
  nrow(shares) == 2L * 3L * ncol(overview_abundance) * length(share_sets),
  all(is.finite(shares$share)),
  all(shares$share >= -1e-10 & shares$share <= 1 + 1e-10)
)
share_totals <- aggregate(share ~ sample + q + type, shares, sum)
stopifnot(nrow(share_totals) == 2L * 3L * ncol(overview_abundance),
          max(abs(share_totals$share - 1)) < 1e-10)
shares$habitat <- sample_metadata$habitat_label[
  match(shares$sample, sample_metadata$sample_id)
]
stopifnot(!anyNA(shares$habitat))
```

## Shares across habitats

The heatmap shows the median share among the 45 samples in each habitat. Its
colour scale is the same in every panel. At neutral $q=1$, a set's share equals
its summed relative abundance. At $q=0$, each detected representative has
equal neutral weight; at $q=2$, abundant representatives carry more weight.
Phylogenetic shares also allocate the weights of tree branches to their
descendant representatives.

```{r share_heatmap_data, include=FALSE}
share_habitat_short <- c(
  "Field soil", "Grassland", "Forest", "Urban green", "Bog/fen",
  "Freshwater", "Saltwater", "Wastewater"
)
stopifnot(length(share_habitat_short) == length(habitat_levels))
share_medians <- aggregate(share ~ habitat + q + type + set, shares, median)
share_medians$Habitat <- factor(
  share_habitat_short[match(share_medians$habitat, habitat_levels)],
  levels = share_habitat_short
)
share_medians$Set <- factor(share_medians$set,
                            levels = rev(names(share_sets)))
share_medians$Type <- factor(share_medians$type,
                             levels = c("neutral", "phylogenetic"),
                             labels = c("Neutral", "Phylogenetic"))
share_medians$Order <- factor(share_medians$q, levels = c(0, 1, 2),
                              labels = c("q = 0", "q = 1", "q = 2"))
stopifnot(!anyNA(share_medians$Habitat),
          nrow(share_medians) == 2L * 3L * length(habitat_levels) *
            length(share_sets))
```

```{r share_heatmap, fig.width=13, fig.height=10, fig.alt="Six heatmaps show the median share of four phyla and all other representatives across eight habitats, for neutral and phylogenetic diversity at orders zero, one, and two."}
ggplot2::ggplot(share_medians,
                ggplot2::aes(x = Habitat, y = Set, fill = share)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.4) +
  ggplot2::scale_fill_gradient(
    low = "#f7f7f7", high = "#8b1e1e", limits = c(0, 1),
    labels = function(x) sprintf("%.0f%%", 100 * x), name = "Median share"
  ) +
  ggplot2::facet_grid(Order ~ Type) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

*Median power-sum shares in each habitat. Medians are taken separately for
each set, so the displayed medians need not sum to one; the five shares sum to
one within every individual sample, type, and order.*

Thermoproteota has a higher $q=1$ share in field soil than in grassland soil.
The next figure shows the sample-level values for both diversity types rather
than only the habitat medians in the heatmap.

```{r share_thermoproteota_data, include=FALSE}
thermoproteota <- shares[
  shares$set == "Thermoproteota" & shares$q == 1 &
    shares$habitat %in% contrast_habitats,
]
thermoproteota$habitat <- factor(thermoproteota$habitat,
                                levels = rev(contrast_habitats))
thermoproteota$Type <- factor(thermoproteota$type,
                              levels = c("neutral", "phylogenetic"),
                              labels = c("Neutral", "Phylogenetic"))
stopifnot(nrow(thermoproteota) == 2L * 45L * length(contrast_habitats))
```

```{r share_thermoproteota, fig.width=10, fig.height=4.5, fig.alt="Boxplots and individual samples compare order-one Thermoproteota power-sum shares in agricultural field and natural grassland soils, with neutral and phylogenetic panels on the same scale."}
ggplot2::ggplot(thermoproteota,
                ggplot2::aes(x = share, y = habitat)) +
  ggplot2::geom_boxplot(outlier.shape = NA, width = 0.55,
                        fill = "#e6c7c7", colour = "#702626") +
  ggplot2::geom_jitter(height = 0.13, width = 0, alpha = 0.4,
                       size = 0.8, colour = "#702626") +
  ggplot2::facet_wrap(~ Type, nrow = 1) +
  ggplot2::scale_x_continuous(
    labels = function(x) sprintf("%.0f%%", 100 * x),
    breaks = seq(0, 0.4, by = 0.1),
    expand = ggplot2::expansion(mult = c(0, 0.03))
  ) +
  ggplot2::labs(x = "Thermoproteota share (q = 1)", y = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
                 strip.text = ggplot2::element_text(face = "bold"))
```

*Each point is one sample; boxes show medians and interquartile ranges. The
panels share an abundance scale. A zero means no representative from this
phylum was detected in that sample.*

The median neutral Thermoproteota share is
`r sprintf("%.1f%%", 100 * median(thermoproteota$share[thermoproteota$type == "neutral" & thermoproteota$habitat == contrast_habitats[1]]))`
in field soil and
`r sprintf("%.1f%%", 100 * median(thermoproteota$share[thermoproteota$type == "neutral" & thermoproteota$habitat == contrast_habitats[2]]))`
in grassland soil. The distributions overlap, and these unpaired samples do
not establish a land-use effect.

## How order and tree weighting change a share

For Pseudomonadota, the next plot compares sample medians and interquartile
ranges across orders in four soil habitats. The lines connect summaries, not
individual samples. A shared vertical scale makes both diversity types
directly visible.

```{r share_order_data, include=FALSE}
share_profile_habitats <- c(
  contrast_habitats, "Urban greenspace soil", "Natural forest soil"
)
pseudomonadota <- data.table::as.data.table(shares[
  shares$set == "Pseudomonadota" &
    shares$habitat %in% share_profile_habitats,
])
share_order <- pseudomonadota[
  , .(median = median(share), lower = as.numeric(quantile(share, 0.25)),
      upper = as.numeric(quantile(share, 0.75))),
  by = .(habitat, type, q)
]
share_order$habitat <- factor(share_order$habitat,
                              levels = share_profile_habitats)
share_order$Type <- factor(share_order$type,
                           levels = c("neutral", "phylogenetic"),
                           labels = c("Neutral", "Phylogenetic"))
stopifnot(nrow(share_order) == length(share_profile_habitats) * 2L * 3L)
```

```{r share_order_plot, fig.width=10, fig.height=7.5, fig.alt="Four panels show Pseudomonadota median power-sum shares and interquartile ranges at orders zero, one, and two, with separate neutral and phylogenetic lines."}
ggplot2::ggplot(share_order,
                ggplot2::aes(x = q, y = median, colour = Type, group = Type)) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = lower, ymax = upper),
                         width = 0.09, linewidth = 0.5) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = c(0, 1, 2)) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1), breaks = seq(0, 1, by = 0.2),
    labels = function(x) sprintf("%.0f%%", 100 * x)
  ) +
  ggplot2::scale_colour_manual(values = c("Neutral" = "#702626",
                                          "Phylogenetic" = "#1b6b75")) +
  ggplot2::facet_wrap(~ habitat, ncol = 2) +
  ggplot2::labs(x = "Diversity order (q)", y = "Pseudomonadota share",
                colour = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
                 strip.text = ggplot2::element_text(face = "bold"))
```

*Points are medians across 45 samples; bars are interquartile ranges. They
describe the selected panel and are not confidence intervals.*

The phylogenetic tree is rooted but not ultrametric. At $q=1$, differences
between neutral and phylogenetic shares reflect root-to-tip path lengths in
addition to abundance. At other orders, shared branch weights also affect
the allocation. The panel has no matched functional-distance matrix, so we
restrict this example to neutral and phylogenetic shares. Habitat and
sampling year are partly confounded; these figures do not establish a habitat
effect.

## Which phyla account for dissimilarity?

We compare the **mean abundance profiles** of agricultural field soil and
natural grassland soil, calculated earlier from 45 samples per habitat. Both
are soils, but the source metadata classify one as agricultural and the other
as natural. This gives the contribution analysis an ecological land-use
question with a shared sample matrix. It is one comparison of two habitat mean
vectors, not the median of the 2,025 individual sample pairs shown in the
[pairwise analysis](#pairwise-beta). The samples are not matched on soil
properties, vegetation, location, or sampling year, so this descriptive
contrast cannot isolate an effect of human land use.

The same five disjoint genome sets used for `hillshare()` cover every
representative. Here `hillcontrib()` attributes the $q=1$ C dissimilarity to
those sets. A contribution is a part of the dissimilarity, whereas the
earlier power-sum share describes one profile on its own.

```{r contrib_inputs, include=FALSE}
contrast_profiles <- habitat_profiles[, contrast_habitats, drop = FALSE]
nitrososphaerales_ids <- genome_metadata$genome_id[
  phylum == "Thermoproteota" &
    genome_metadata$order == "o__Nitrososphaerales"
]
nitrososphaerales_fraction <- colSums(
  contrast_profiles[nitrososphaerales_ids, , drop = FALSE]
) / colSums(contrast_profiles[share_sets$Thermoproteota, , drop = FALSE])
stopifnot(length(nitrososphaerales_ids) > 0L,
          all(nitrososphaerales_fraction > 0.99))
```

```{r share_hillcontrib, message=FALSE, class.source='script-source hilldiv3-source'}
contrib_neutral <- hilldiv3::hillcontrib(
  contrast_profiles, q = 1, metric = "C", type = "neutral",
  sets = share_sets
)
contrib_phylogenetic <- hilldiv3::hillcontrib(
  contrast_profiles, q = 1, metric = "C", tree = genome_tree,
  type = "phylogenetic", sets = share_sets
)
```

```{r contrib_checks, include=FALSE, message=FALSE}
contrast_neutral <- hilldiv3::hillpair(
  contrast_profiles, q = 1, metric = "C", type = "neutral", out = "tibble"
)
contrast_phylogenetic <- hilldiv3::hillpair(
  contrast_profiles, q = 1, metric = "C", tree = genome_tree,
  type = "phylogenetic", out = "tibble"
)
stopifnot(
  abs(sum(contrib_neutral$contribution) - contrast_neutral$value) < 1e-10,
  abs(sum(contrib_phylogenetic$contribution) -
        contrast_phylogenetic$value) < 1e-10,
  abs(sum(contrib_neutral$share) - 1) < 1e-10,
  abs(sum(contrib_phylogenetic$share) - 1) < 1e-10
)
contributions <- rbind(
  transform(contrib_neutral, Type = "Neutral"),
  transform(contrib_phylogenetic, Type = "Phylogenetic")
)
contributions$Type <- factor(contributions$Type,
                             levels = c("Neutral", "Phylogenetic"))
contributions$set <- factor(contributions$set, levels = names(share_sets))
```

```{r contrib_plot, fig.width=9, fig.height=4.2, fig.alt="Two stacked bars partition order-one C dissimilarity between mean agricultural field and natural grassland soil profiles across five representative sets. Neutral dissimilarity is about 0.29 and phylogenetic dissimilarity about 0.05. Actinomycetota contributes the most to both values."}
ggplot2::ggplot(contributions,
                ggplot2::aes(x = Type, y = contribution, fill = set)) +
  ggplot2::geom_col(width = 0.55) +
  ggplot2::coord_flip() +
  ggplot2::scale_y_continuous(
    limits = c(0, 0.3), breaks = seq(0, 0.3, by = 0.1),
    expand = ggplot2::expansion(mult = c(0, 0))
  ) +
  ggplot2::scale_fill_manual(values = c(
    Pseudomonadota = "#702626", Actinomycetota = "#ca725c",
    Acidobacteriota = "#d8b763", Thermoproteota = "#1b6b75",
    Other = "#a6a6a6"
  )) +
  ggplot2::labs(x = NULL, y = "Contribution to C dissimilarity (q = 1)",
                fill = "Representative set") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())
```

*The bar lengths are the C dissimilarities for the same two mean profiles;
their segments sum to those values. They are not median contributions among
sample pairs.*

Neutral C dissimilarity is
`r sprintf("%.3f", contrast_neutral$value)` and phylogenetic C dissimilarity is
`r sprintf("%.3f", contrast_phylogenetic$value)`. Actinomycetota accounts for
`r sprintf("%.1f%%", 100 * contrib_neutral$share[contrib_neutral$set == "Actinomycetota"])`
of the neutral value and
`r sprintf("%.1f%%", 100 * contrib_phylogenetic$share[contrib_phylogenetic$set == "Actinomycetota"])`
of the phylogenetic value. Pseudomonadota and Thermoproteota also make
substantial contributions. Phylogenetic attribution allocates shared branch
turnover to descendant representatives under `hillcontrib()`'s
specified rule; it is not a unique biological assignment of each branch to a
phylum.

### How phylogenetic weighting redistributes contributions

Thermoproteota accounts for
`r sprintf("%.1f%%", 100 * contrib_neutral$share[contrib_neutral$set == "Thermoproteota"])`
of the neutral C value and
`r sprintf("%.1f%%", 100 * contrib_phylogenetic$share[contrib_phylogenetic$set == "Thermoproteota"])`
of the phylogenetic value. Pseudomonadota moves in the opposite direction,
from
`r sprintf("%.1f%%", 100 * contrib_neutral$share[contrib_neutral$set == "Pseudomonadota"])`
to
`r sprintf("%.1f%%", 100 * contrib_phylogenetic$share[contrib_phylogenetic$set == "Pseudomonadota"])`
of the phylogenetic value. Representatives assigned to Nitrososphaerales make
up
`r sprintf("%.1f%%", 100 * nitrososphaerales_fraction[1])`
and
`r sprintf("%.1f%%", 100 * nitrososphaerales_fraction[2])`
of Thermoproteota abundance in the field and grassland mean profiles,
respectively. That is a taxonomic assignment, not a measurement of function.
The contribution shares describe two different C values; no significance test
is involved. The figure shows one small clade from each phylum on the same
branch-gap colour scale. The nonlinear colour mapping keeps smaller nonzero
gaps visible; the legend reports the original gap values.

```{r branch_contrast_data, include=FALSE}
source("scripts/plot-branch-contrast.R")
branch_contrast <- branch_contrast_example(
  contrast_profiles, genome_tree, share_sets,
  contrib_neutral, contrib_phylogenetic
)
stopifnot(
  abs(sum(branch_contrast$branch_gap) / log(2) -
        contrast_phylogenetic$value) < 1e-10,
  sum(branch_contrast$branch_gap[
    branch_contrast$focus_edges$Thermoproteota
  ]) > 3 * sum(branch_contrast$branch_gap[
    branch_contrast$focus_edges$Pseudomonadota
  ])
)
dir.create("figures", showWarnings = FALSE)
ggplot2::ggsave("figures/branch-contrast.png",
                plot = branch_contrast$plot, width = 10.5, height = 7.3,
                dpi = 192)
```

```{r branch_contrast_plot, echo=FALSE, fig.width=10.5, fig.height=7.3, fig.alt="Four-tip clades from Thermoproteota and Pseudomonadota on the same branch-gap colour scale for the agricultural field versus natural grassland soil comparison. The selected Thermoproteota clade, assigned to Nitrososphaerales, has a larger combined branch gap than the selected Pseudomonadota clade. Field and grassland clade totals and phylum-wide contribution shares are shown."}
print(branch_contrast$plot)
```

*Each panel shows a separate four-MAG clade in the source tree. Horizontal
branches are drawn to scale and coloured by the same length-weighted $q=1$
abundance-gap scale, with nonlinear colour spacing; grey vertical lines
connect them. All four Thermoproteota representatives shown are assigned to
Nitrososphaerales, and their clade has a larger combined branch gap than the
selected Pseudomonadota clade. The shares in the panel headings come from all
representatives and branches in each
phylum; these selected clades illustrate the contrast but do not alone
determine the phylum-wide percentages.*
