---
title: "AlberdiLab | hilldiv3 manuscript"
subtitle: "Microflora Danica worked example: data and diversity"
author: "Alberdi Lab"
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/hilldiv3-ms/
description: |
  Data and analysis code for the Microflora Danica worked example in the hilldiv3 manuscript.
link-citations: yes
github-repo: alberdilab/hilldiv3-ms
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
  class.source = "script-source",
  class.output = "script-output",
  comment = NA
)
```

# Introduction

This webbook contains the data and reproducible code for the Microflora Danica
worked example in the hilldiv3 manuscript. The selected panel contains sample
and genome metadata, a representative MAG abundance matrix, and a phylogenetic
tree. The chapters describe the panel and begin neutral and phylogenetic
diversity analyses.

## Prepare the R environment

### Environment

Clone the [GitHub repository](https://github.com/alberdilab/hilldiv3-ms) and
open `hilldiv3-ms.Rproj`, or create a new RStudio project from:

```text
https://github.com/alberdilab/hilldiv3-ms.git
```

### Libraries

The book uses these R packages:

```{r load_libraries, message=FALSE}
library(data.table)
library(ape)
library(knitr)
library(ggplot2)
library(maps)
library(hilldiv3)
```

## Analysis steps

### Step 0 - Data preparation

The [data preparation](#data-preparation) chapter describes the committed
Microflora Danica panel, loads it, and checks its sample, genome, abundance,
and tree alignment. The source-download and preparation scripts are in
`scripts/`.

### Step 1 - Data overview

The [data overview](#data-overview) chapter summarizes sample coverage,
sampling dates, geography, and representative genomes.

### Step 2 - Initial diversity analyses

The [diversity analyses](#diversity) chapter calculates neutral and
phylogenetic Hill numbers at orders 0, 1, and 2, then partitions diversity
within and between the eight habitat classes using `hilldiv3`.

### Step 3 - Pairwise beta diversity

The [pairwise beta diversity](#pairwise-beta) chapter compares every pair of
samples with neutral and phylogenetic Hill-number C dissimilarity. It summarizes
within- and between-habitat pairs and shows habitat heatmaps and sample-level
ordinations at $q=0$ and $q=1$.
